---
title: "Examples"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article summarizes examples drawn from StackOverflow and
elsewhere. Start by loading the [rjsoncons][] and [dplyr][] packages

```{r rjsoncons, message = FALSE}
library(rjsoncons)
library(dplyr)
```

[rjsoncons]: https://mtmorgan.github.io/rjsoncons
[dplyr]: https://CRAN.R-project.org/package=dplyr

# Query and pivot

## Data frame columns as NDJSON

<https://stackoverflow.com/questions/76447100>

This question presented a `tibble` where one column contained JSON expressions.

```{r, echo = FALSE}
df <- structure(list(
    subject = c("dtv85251vucquc45", "mcj8vdqz7sxmjcr0"),
    response = c(
        "{\"P0_Q0\":{\"aktiv\":2,\"bekümmert\":3,\"interessiert\":4,\"freudig erregt\":2,\"verärgert\":2,\"stark\":0,\"schuldig\":1,\"erschrocken\":1,\"feindselig\":1,\"angeregt\":2},\"P1_Q0\":{\"stolz\":1,\"gereizt\":1,\"begeistert\":2,\"beschämt\":2,\"wach\":1,\"nervös\":1,\"entschlossen\":1,\"ängstlich\":1,\"aufmerksam\":2,\"durcheinander\":2}}", 
        "{\"P0_Q0\":{\"aktiv\":1,\"bekümmert\":3,\"interessiert\":1,\"freudig erregt\":1,\"verärgert\":0,\"stark\":0,\"schuldig\":2,\"erschrocken\":0,\"feindselig\":0,\"angeregt\":1},\"P1_Q0\":{\"stolz\":2,\"gereizt\":0,\"begeistert\":1,\"beschämt\":0,\"wach\":2,\"nervös\":0,\"entschlossen\":0,\"aufmerksam\":2,\"durcheinander\":0,\"ängstlich\":0}}"
    )),
    class = c("tbl_df", "tbl", "data.frame"), row.names = 1:2)
```

```{r}
df
```

The goal was to extract the fields of each `P0_Q0` element as new
columns. The `response` column can be viewed as NDJSON, so we can use
`pivot(df$response, "P0_Q0")` for the hard work, and `bind_cols()` to
prepend subject

```{r}
bind_cols(
    df |> select(subject),
    df |> pull(response) |> j_pivot("P0_Q0", as = "tibble")
)
```

My [initial response][76447258/547331] was early in package
development, and motivated `j_pivot()` as an easier way to perform the
common operation of transforming a JSON array-of-objects to and *R*
data frame.

[76447258/547331]: https://stackoverflow.com/a/76447258/547331

## Constructing a pivot object using [JMESPath][]

<https://stackoverflow.com/questions/78029215>

This question had an array of objects, each of which had a single
unique key-value pair. 

```{r key-value}
json <- '{
    "encrypted_values":[
        {"name_a":"value_a"}, 
        {"name_b":"value_b"}, 
        {"name_c":"value_c"}
    ]
}'
```

The goal was to create a tibble with keys in one column, and values in
another.  `jsonlite::fromJSON()` or `j_pivot(json,
"encrypted_values")` simplify the result to a tibble with a column for
each object key, which is not desired.

```{r key-value-wrong}
jsonlite::fromJSON(json)$encrypted_values
j_pivot(json, "encrypted_values", as = "tibble")
```

Instead, write a JMESPath query that extracts an object with the keys
as one element, and values as another. This uses `@` to represent the
current mode, and `keys()` and `values()` functions to extract
associated elements. The trailing `[]` converts an array-of-arrays of
keys (for example) to a simple array of keys.

```{r key-value-query}
query <- '{
    name : encrypted_values[].keys(@)[],
    value: encrypted_values[].values(@)[]
}'
j_pivot(json, query, as = "tibble")
```

[JMESPath]: https://jmespath.org

## Reading from URLs

<https://stackoverflow.com/questions/78023560>

This question illustrates [rjsoncons][] ability to read URLs; the
query itself extracts from the `fixtures` array of objects specific
nested elements, and is similar to the previous question. In practice,
I used `json <- readLines(url)` to create a local copy of the data to
use while developing the query.

```{r}
url <- "https://www.nrl.com/draw//data?competition=111&season=2024"
query <- 'fixtures[].{
    homeTeam: homeTeam.nickName, 
    awayTeam: awayTeam.nickName
}'
j_pivot(url, query, as = "tibble")
```

The easiest path to a more general answer (extract all members of
'homeTeam' and 'awayTeam' as a tibble) might, like the posted answer,
combine JSON extraction and [tidyr][].

```{r}
query <- 'fixtures[].{ homeTeam: homeTeam, awayTeam: awayTeam }'
j_pivot(url, query, as = "tibble") |>
    tidyr::unnest_wider(c("homeTeam", "awayTeam"), names_sep = "_")
```

[tidyr]: https://CRAN.R-project.org/package=tidyr

## Deeply nested objects

<https://stackoverflow.com/questions/77998013> 

The details of this question are on StackOverflow, and the following
code chunks are not evaluated directly. The example has several
interesting elements:

- The JSON is quite large (about 90 Mb), so processing is not
  immediate. While developing the query, I focused on a subset of the
  data for a more interactive experience.

    ```{r, eval = FALSE}
    Crime2013 <- j_query(json, "x.calls[9].args")
    ```

- JSON array indexing is 0-based, in contrast to 1-based *R* indexing.
- In developing the JSON query, I spent quite a bit of time viewing
  results using `listviewer::jsonedit()`, e.g.,

    ```{r, eval = FALSE}
    listviewer::jsonedit(j_query(Crime2013, "[0][*]"))
    ```

The objects of interest are polygon coordinates nested deeply in the
JSON, at the location implied by JMESPath. One polygon is at

```{r, eval = FALSE}
query <- "x.calls[9].args[0][0][0][0]"
j_pivot(json, query, as = "tibble")
## # A tibble: 27 × 2
##      lng   lat
##    <dbl> <dbl>
##  1 -43.3 -22.9
##  2 -43.3 -22.9
##  3 -43.3 -22.9
##  4 -43.3 -22.9
##  5 -43.3 -22.9
##  6 -43.3 -22.9
##  7 -43.3 -22.9
##  8 -43.3 -22.9
##  9 -43.3 -22.9
## 10 -43.3 -22.9
## # ℹ 17 more rows
## # ℹ Use `print(n = ...)` to see more rows
```

There are 3618 of these polygons, and they are extracted by using a
wild-card `*` in place of an index `0` at a particular place in the
path.

```{r, eval = FALSE}
query <- "x.calls[9].args[0][*][0][0]"
j_pivot(json, query, as = "tibble")
## # A tibble: 3,618 × 2
##    lng        lat       
##    <list>     <list>    
##  1 <dbl [27]> <dbl [27]>
##  2 <dbl [6]>  <dbl [6]> 
##  3 <dbl [5]>  <dbl [5]> 
##  4 <dbl [42]> <dbl [42]>
##  5 <dbl [6]>  <dbl [6]> 
##  6 <dbl [8]>  <dbl [8]> 
##  7 <dbl [4]>  <dbl [4]> 
##  8 <dbl [6]>  <dbl [6]> 
##  9 <dbl [13]> <dbl [13]>
## 10 <dbl [12]> <dbl [12]>
## # ℹ 3,608 more rows
## # ℹ Use `print(n = ...)` to see more rows
```

`tidyr::unnest()` could be used to create a 'long' version of this
result, as illustrated in [my response][78035500/547331].

While exploring the data, the JMESPath function `length()` suggested
two anomalies in the data (not all paths in `Crime2013` have just one
polygon; a path used to identify polygons has only 2022 elements, but
there are 3618 polygons); this could well be a misunderstanding on my
part.

[78035500/547331]: https://stackoverflow.com/a/78035500/547331

# Patch

## Moving elements

<https://stackoverflow.com/questions/78047988>

The example is not completely reproducible, but the challenge is that
the [igraph][] package produces JSON like

```{r}
data <- '
{
  "nodes": [
     {
        "name": "something"
     },
     {
        "name": "something_else"
     }
  ],
  "links": [
      {
         "source": "something",
         "target": "something_else"
      }
   ],
  "attributes": {
      "directed": false
  }
}'
```

but the desired data moves the 'directed' attribute to a top-level
field. From the [JSON patch][] documentation, the patch is a single
'move' operation from `/attributes/directed` to the top-level `/`:

```{r}
patch <- '[
    {"op": "move", "from": "/attributes/directed", "path": "/directed"}
]'
```

The patch is accomplished with

```{r}
patched_data <- j_patch_apply(data, patch)
```

This JSON string could be visualized with
`listviwer::jsonedit(patched_data)`, or `patched_data |> as_r() |>
str()`, or

```{r}
patched_data |>
    jsonlite::prettify()
```

[JSON patch]: https://jsonpatch.com

# Session information {.unnumbered}

```{r}
sessionInfo()
```
