---
title: "Examples"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article summarizes examples drawn from StackOverflow and
elsewhere. Start by loading the [rjsoncons][] and [dplyr][] packages

```{r rjsoncons, message = FALSE}
library(rjsoncons)
library(dplyr)
```

[rjsoncons]: https://mtmorgan.github.io/rjsoncons
[dplyr]: https://CRAN.R-project.org/package=dplyr

# Query and pivot

## Data frame columns as NDJSON

<https://stackoverflow.com/questions/76447100>

This question presented a `tibble` where one column contained JSON expressions.

```{r, echo = FALSE}
df <- structure(list(
    subject = c("dtv85251vucquc45", "mcj8vdqz7sxmjcr0"),
    response = c(
        "{\"P0_Q0\":{\"aktiv\":2,\"bekümmert\":3,\"interessiert\":4,\"freudig erregt\":2,\"verärgert\":2,\"stark\":0,\"schuldig\":1,\"erschrocken\":1,\"feindselig\":1,\"angeregt\":2},\"P1_Q0\":{\"stolz\":1,\"gereizt\":1,\"begeistert\":2,\"beschämt\":2,\"wach\":1,\"nervös\":1,\"entschlossen\":1,\"ängstlich\":1,\"aufmerksam\":2,\"durcheinander\":2}}", 
        "{\"P0_Q0\":{\"aktiv\":1,\"bekümmert\":3,\"interessiert\":1,\"freudig erregt\":1,\"verärgert\":0,\"stark\":0,\"schuldig\":2,\"erschrocken\":0,\"feindselig\":0,\"angeregt\":1},\"P1_Q0\":{\"stolz\":2,\"gereizt\":0,\"begeistert\":1,\"beschämt\":0,\"wach\":2,\"nervös\":0,\"entschlossen\":0,\"aufmerksam\":2,\"durcheinander\":0,\"ängstlich\":0}}"
    )),
    class = c("tbl_df", "tbl", "data.frame"), row.names = 1:2)
```

```{r}
df
```

The goal was to extract the fields of each `P0_Q0` element as new
columns. The `response` column can be viewed as NDJSON, so we can use
`pivot(df$response, "P0_Q0")` for the hard work, and `bind_cols()` to
prepend subject

```{r}
bind_cols(
    df |> select(subject),
    df |> pull(response) |> j_pivot("P0_Q0", as = "tibble")
)
```

My [initial response][76447258/547331] was early in package
development, and motivated `j_pivot()` as an easier way to perform the
common operation of transforming a JSON array-of-objects to and *R*
data frame.

[76447258/547331]: https://stackoverflow.com/a/76447258/547331

## Constructing a pivot object using [JMESPath][]

<https://stackoverflow.com/questions/78029215>

This question had an array of objects, each of which had a single
unique key-value pair. 

```{r key-value}
json <- '{
    "encrypted_values":[
        {"name_a":"value_a"}, 
        {"name_b":"value_b"}, 
        {"name_c":"value_c"}
    ]
}'
```

The goal was to create a tibble with keys in one column, and values in
another.  `jsonlite::fromJSON()` or `j_pivot(json,
"encrypted_values")` simplify the result to a tibble with a column for
each object key, which is not desired.

```{r key-value-wrong}
jsonlite::fromJSON(json)$encrypted_values
j_pivot(json, "encrypted_values", as = "tibble")
```

Instead, write a JMESPath query that extracts an object with the keys
as one element, and values as another. This uses `@` to represent the
current mode, and `keys()` and `values()` functions to extract
associated elements. The trailing `[]` converts an array-of-arrays of
keys (for example) to a simple array of keys.

```{r key-value-query}
query <- '{
    name : encrypted_values[].keys(@)[],
    value: encrypted_values[].values(@)[]
}'
j_pivot(json, query, as = "tibble")
```

[JMESPath]: https://jmespath.org

## Constructing a pivot object using [JMESPath][]: a second example

<https://stackoverflow.com/questions/78727724>

The question asks about creating a tibble from a complex JSON data
structure. Here is a reproducible example to retrieve JSON;
unfortunately the host does not resolve when run as a GitHub action,
so the example here is not fully evaluated.

```{r, eval = FALSE}
start_date <- end_date <- Sys.Date()

res <- httr::GET(
       url = "https://itc.aeso.ca/itc/public/api/v2/interchange",
       query = list(
           beginDate = format(start_date, "%Y%m%d"),
           endDate = format(end_date, "%Y%m%d"),
           Accept = "application/json"
       )
   )
json <- httr::content(res, as = "text", encoding = "UTF-8")
```

Explore the JSON using [listviewer][].

```{r, eval = FALSE}
listviewer::jsonedit(json)
```

Write a query that extracts, directly from the JSON, some of the
fields of interest. The queries are written using [JMESPath][]. A
simple example extracts the 'date' from the path
`return.BcIntertie.Allocations[].date`

```{r, eval = FALSE}
path <- 'return.BcIntertie.Allocations[].date'
j_query(json, path) |>
    str()
## chr "[\"2024-07-10\",\"2024-07-10\",\"2024-07-10\",\"2024-07-10\",\"2024-07-10\",\"2024-07-10\",\"2024-07-10\",\"202"| __truncated__
```

Expand on this by querying several different fields, and then
re-formating the query into a new JSON object.  Develop the code by
querying / viewing until things look like a JSON array-of-objects

```{r, eval = FALSE}
path <- paste0(
    'return.{',
        'date: BcIntertie.Allocations[].date,',
        'he: BcIntertie.Allocations[].he,',
        'bc_import: BcIntertie.Allocations[].import.atc,',
        'bc_export:  BcIntertie.Allocations[].export.atc,',
        'matl_import: MatlIntertie.Allocations[].import.atc,',
        'matl_export:  MatlIntertie.Allocations[].export.atc',
    '}'
)
```

```{r, eval = FALSE}
j_query(json, path) |>
    listviewer::jsonedit()
```

Finally, run `j_pivot()` to transform the JSON to a tibble.

```{r, eval = FALSE}
j_pivot(json, path, as = "tibble")
```

The result is

```
## # A tibble: 48 × 6
##    date       he    bc_import bc_export matl_import matl_export
##    <chr>      <chr>     <int>     <int>       <int>       <int>
##  1 2024-07-10 4           750       950         295         300
##  2 2024-07-10 5           750       950         295         300
##  3 2024-07-10 6           750       950         295         300
##  4 2024-07-10 7           750       950         295         300
##  5 2024-07-10 8           750       950         295         300
##  6 2024-07-10 9           750       950         295         300
##  7 2024-07-10 10          750       950         295         300
##  8 2024-07-10 11          750       950         295         300
##  9 2024-07-10 12          750       950         295         300
## 10 2024-07-10 13          750       950         295         300
## # ℹ 38 more rows
## # ℹ Use `print(n = ...)` to see more rows
```

## Reshaping nested records

<https://stackoverflow.com/questions/78952424>

This question wants to transform JSON into a [data.table][]. A
previous answer uses `rbindlist()` (similar to `dplyr::bind_rows()`)
to transform structured lists to data.tables. Here is the sample data

```{r}
json <- '[
  {
    "version_id": "123456",
    "data": [
      {
        "review_id": "1",
        "rating": 5,
        "review": "This app is great",
        "date": "2024-09-01"
      },
      {
        "review_id": "2",
        "rating": 1,
        "review": "This app is terrible",
        "date": "2024-09-01"
      }
    ]
  },
  {
    "version_id": "789101",
    "data": [
      {
        "review_id": "3",
        "rating": 3,
        "review": "This app is OK",
        "date": "2024-09-01"
      }
    ]
  }
]'
```

The desired data.table is flattened to include `version_id` and each
field of `data[]` as columns in the table, with the complication that
`version_id` needs to be replicated for each element of `data[]`.

The rjsoncons [answer][78952424/547331] illustrates several
approaches. The approach most cleanly separating data transformation
and data.table construction using [JMESPath][] creates an array of
objects where each `version_id` is associated with *vectors* of
`review_id`, etc., corresponding to that version.

```{r}
query <-
    "[].{
        version_id: version_id,
        review_id: data[].review_id,
        rating: data[].rating,
        review: data[].review,
        date: data[].date
    }"
```

As an *R* object, this is exactly handled by `rbindlist()`. Note that
a pivot is not involved.

```{r}
records <- j_query(json, query, as = "R")
data.table::rbindlist(records)
```

[dplyr][]'s `bind_rows()` behaves similarly:

```{r}
dplyr::bind_rows(records)
```

[data.table]: https://CRAN.R-project.org/package=data.table
[78952424/547331]: https://stackoverflow.com/a/78967138/547331

## Reading from URLs

<https://stackoverflow.com/questions/78023560>

This question illustrates [rjsoncons][] ability to read URLs; the
query itself extracts from the `fixtures` array of objects specific
nested elements, and is similar to the previous question. In practice,
I used `json <- readLines(url)` to create a local copy of the data to
use while developing the query.

```{r}
url <- "https://www.nrl.com/draw//data?competition=111&season=2024"
query <- 'fixtures[].{
    homeTeam: homeTeam.nickName, 
    awayTeam: awayTeam.nickName
}'
j_pivot(url, query, as = "tibble")
```

The easiest path to a more general answer (extract all members of
'homeTeam' and 'awayTeam' as a tibble) might, like the posted answer,
combine JSON extraction and [tidyr][].

```{r}
query <- 'fixtures[].{ homeTeam: homeTeam, awayTeam: awayTeam }'
j_pivot(url, query, as = "tibble") |>
    tidyr::unnest_wider(c("homeTeam", "awayTeam"), names_sep = "_")
```

[tidyr]: https://CRAN.R-project.org/package=tidyr

## Deeply nested objects

<https://stackoverflow.com/questions/77998013> 

The details of this question are on StackOverflow, and the following
code chunks are not evaluated directly. The example has several
interesting elements:

- The JSON is quite large (about 90 Mb), so processing is not
  immediate. While developing the query, I focused on a subset of the
  data for a more interactive experience.

    ```{r, eval = FALSE}
    Crime2013 <- j_query(json, "x.calls[9].args")
    ```

- JSON array indexing is 0-based, in contrast to 1-based *R* indexing.
- In developing the JSON query, I spent quite a bit of time viewing
  results using `listviewer::jsonedit()`, e.g.,

    ```{r, eval = FALSE}
    listviewer::jsonedit(j_query(Crime2013, "[0][*]"))
    ```

The objects of interest are polygon coordinates nested deeply in the
JSON, at the location implied by JMESPath. One polygon is at

```{r, eval = FALSE}
query <- "x.calls[9].args[0][0][0][0]"
j_pivot(json, query, as = "tibble")
## # A tibble: 27 × 2
##      lng   lat
##    <dbl> <dbl>
##  1 -43.3 -22.9
##  2 -43.3 -22.9
##  3 -43.3 -22.9
##  4 -43.3 -22.9
##  5 -43.3 -22.9
##  6 -43.3 -22.9
##  7 -43.3 -22.9
##  8 -43.3 -22.9
##  9 -43.3 -22.9
## 10 -43.3 -22.9
## # ℹ 17 more rows
## # ℹ Use `print(n = ...)` to see more rows
```

There are 3618 of these polygons, and they are extracted by using a
wild-card `*` in place of an index `0` at a particular place in the
path.

```{r, eval = FALSE}
query <- "x.calls[9].args[0][*][0][0]"
j_pivot(json, query, as = "tibble")
## # A tibble: 3,618 × 2
##    lng        lat       
##    <list>     <list>    
##  1 <dbl [27]> <dbl [27]>
##  2 <dbl [6]>  <dbl [6]> 
##  3 <dbl [5]>  <dbl [5]> 
##  4 <dbl [42]> <dbl [42]>
##  5 <dbl [6]>  <dbl [6]> 
##  6 <dbl [8]>  <dbl [8]> 
##  7 <dbl [4]>  <dbl [4]> 
##  8 <dbl [6]>  <dbl [6]> 
##  9 <dbl [13]> <dbl [13]>
## 10 <dbl [12]> <dbl [12]>
## # ℹ 3,608 more rows
## # ℹ Use `print(n = ...)` to see more rows
```

`tidyr::unnest()` could be used to create a 'long' version of this
result, as illustrated in [my response][78035500/547331].

While exploring the data, the JMESPath function `length()` suggested
two anomalies in the data (not all paths in `Crime2013` have just one
polygon; a path used to identify polygons has only 2022 elements, but
there are 3618 polygons); this could well be a misunderstanding on my
part.

[78035500/547331]: https://stackoverflow.com/a/78035500/547331

## JSONPath wildcards

<https://stackoverflow.com/questions/78029215>

This question retrieves a JSON representation of the hierarchy of
departments within German research institutions.  The interest is in
finding the path between two departments.

The answer posted on StackOverflow translates the JSON list-of-lists
structure describing the hierarchy into an *R* list-of-lists and uses
a series of complicated manipulations to form a tibble suitable for
querying.

The approach here recognizes this as a graph-based problem. The goal
is to construct a graph from JSON, and then use graph algorithms to
identify the shortest path between nodes.

I followed @margusl to extract JSON from the web page.

```{r}
library(rvest)
html <- read_html("https://www.gerit.org/en/institutiondetail/10282")

## scrape JSON
xpath <- '//script[contains(text(),"window.__PRELOADED_STATE__")]'
json <- 
  html |>
  html_element(xpath = xpath) |>
  html_text() |>
  sub(pattern = "window.__PRELOADED_STATE__ = ", replacement = "", fixed = TRUE)
```

I then used [listviewer][] to explore the JSON interactively, to get a
feel for the structure.

```{r, eval = FALSE}
listviewer::jsonedit(json)
```

[rvest]: https://rvest.tidyverse.org/
[listviewer]: https://cran.r-project.org/package=listviewer

Obtain the root of the institutional hierarchy with a simple path
traversal, using [JSONPath][] syntax. JSONPath provides 'wild-card'
syntax, which is convenient when querying hierarchical data
representations.

```{r, message = FALSE}
library(rjsoncons)
library(dplyr)
tree <- j_query(json, "$.institutionDetail.institution.tree")
```

[rjsoncons]: https://mtmorgan.github.io/rjsoncons
[JSONPath]: https://goessner.net/articles/JsonPath/

It looks like each institution is a (directed, acyclic) graph, with
nodes representing each division in the institution. 

The nodes are easy to reconstruct. Look for keys `id`, `name.de`, and
`name.en`using the wild-card `..`, meaning 'at any depth'. Each of
these queries returns a vector. Use them to construct a
tibble. Although the `id` key is an integer in the JSON, it seems
appropriate to think of these as character-valued.

```{r}
## these are all nodes, including the query node; each node has an
## 'id' and german ('de') and english ('en') name.
nodes <- tibble(
    ## all keys 'id' , 'name.de', and 'name.en', under 'tree' with
    ## wild-card '..' matching
    id =
        j_query(tree, "$..id", as = "R") |>
        as.character(),
    de = j_query(tree, "$..name.de", as = "R"),
    en = j_query(tree, "$..name.en", as = "R")
)
```

There are `r NROW(nodes)` nodes in this institution.

```{r}
nodes
```

The edges are a little tricky to reconstruct from the nested structure
in the JSON. Start with the id and number of children of each node.

```{r}
id <-
    j_query(tree, "$..id", as = "R") |>
    as.character()
children_per_node <- j_query(tree, "$..children.length", as = "R")
```

I developed the `edgelist()` function (in the folded code chunk below)
to transform `id` and `children_per_node` into a two-column matrix of
from-to relations. In the function, `parent_id` and `n_children` are
stacks used to capture the hierarchical structure of the data. `level`
represents the current level in the hierarchy. The algorithm walks
along the `id` input, records the from/to relationship implied by `n`,
and then completes that level of the hierarchy (if `n_children[level]
== 0`), pushes the next level onto the stack, or continues to the next
`id`.

```{r}
edgelist <- function(id, n) {
    stopifnot(identical(length(id), length(n)))
    parent_id <- n_children <- integer()
    from <- to <- integer(length(id) - 1L)
    level <- 0L
    for (i in seq_along(id)) {
        if (i > 1) {
            ## record link from parent to child
            from[i - 1] <- tail(parent_id, 1L)
            to[i - 1] <- id[i]
            n_children[level] <- n_children[level] - 1L
        }
        if (level > 0 && n_children[level] == 0L) {
            ## 'pop' level
            level <- level - 1L
            parent_id <- head(parent_id, -1L)
            n_children <- head(n_children, -1L)
        }
        if (n[i] != 0) {
            ## 'push' level
            level <- level + 1L
            parent_id <- c(parent_id, id[i])
            n_children <- c(n_children, n[i])
        }
    }
    tibble(from, to)
}
```

The matrix of edges, with columns `from` and `to`, is then

```{r}
edges <- edgelist(id, children_per_node)
edges
```

The computation is tricky, but not too inefficient. There are only two
queries of the JSON object (for `id` and `children_per_node`) and the
`R` iteration over elements of `id` is not too extensive.

I used the [tidygraph][] package to represent the graph from the nodes
and edges.

```{r, message = FALSE}
tg <- tidygraph::tbl_graph(nodes = nodes, edges = edges, node_key = "id")
tg
```

I then used `convert()` to find the graph with the shortest path
between two nodes, and extracted the tibble of nodes.

```{r}
tg |>
    ## find the shortest path between two nodes...
    tidygraph::convert(
        tidygraph::to_shortest_path,
        de == "Fakultät 1: Wirtschafts- und Sozialwissenschaftliche Fakultät",
        de == "Professur Hölzl"
    ) |>
    ## extract the nodes from the resulting graph
    as_tibble("nodes") |>
    select(id, de)
```

This is the answer to the question posed in the StackOverflow post.

It can be fun to try and visualize the graph, e.g., using [ggraph][]

```{r}
ggraph::ggraph(tg, "tree", circular = TRUE) +
    ggraph::geom_edge_elbow()
```

[tidygraph]: https://tidygraph.data-imaginist.com
[ggraph]: https://ggraph.data-imaginist.com


# Patch

## Moving elements

<https://stackoverflow.com/questions/78047988>

The example is not completely reproducible, but the challenge is that
the [igraph][] package produces JSON like

```{r}
data <- '
{
  "nodes": [
     {
        "name": "something"
     },
     {
        "name": "something_else"
     }
  ],
  "links": [
      {
         "source": "something",
         "target": "something_else"
      }
   ],
  "attributes": {
      "directed": false
  }
}'
```

but the desired data moves the 'directed' attribute to a top-level
field. From the [JSON patch][] documentation, the patch is a single
'move' operation from `/attributes/directed` to the top-level `/`:

```{r}
patch <- '[
    {"op": "move", "from": "/attributes/directed", "path": "/directed"}
]'
```

The patch is accomplished with

```{r}
patched_data <- j_patch_apply(data, patch)
```

This JSON string could be visualized with
`listviwer::jsonedit(patched_data)`, or `patched_data |> as_r() |>
str()`, or

```{r}
patched_data |>
    jsonlite::prettify()
```

[JSON patch]: https://jsonpatch.com
[igraph]: https://CRAN.R-project.org/package=igraph

# Session information {.unnumbered}

```{r}
sessionInfo()
```
